---
title: "Airbnb vs Long-Term Rentals in Christchurch"
subtitle: "Deliverable 5 questions, October 2025 - June 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(readr)
```

```{r}
# made by join_listings_bonds.Rmd, so knit that first
joined <- read_csv("Christchurch_with_bonds.csv",
                   col_types = cols(id = col_character(),
                                    host_id = col_character(),
                                    area_code = col_character(),
                                    .default = col_guess()))
```

# 1. Median Airbnb price in Christchurch Central (326600)

```{r}
central <- joined %>% filter(area_code == "326600")

nrow(central)             # listing-months
n_distinct(central$id)    # listings
```

```{r}
# observed prices only: Dec-Feb have no prices, and prices over $2,000 are errors
observed <- central %>% filter(price_was_imputed == FALSE)

nrow(observed)
median(observed$price)
```

```{r}
# check 1: include the imputed prices
median(central$price_imputed)

# check 2: count each listing once
per_listing <- observed %>%
  group_by(id) %>%
  summarise(price = median(price))

median(per_listing$price)

# for comparison: all of Christchurch
median(joined$price[joined$price_was_imputed == FALSE])
```

## Result: $238 per night

| Median price per night | |
|---|---|
| Christchurch Central (326600) | **$238** |
| All of Christchurch | $205 |
| Central costs more by | $33 (16%) |

- Based on 705 real prices from 150 listings.
- Left out: December to February (no prices) and 8 prices over $2,000.
- Counting other ways gives $236 and $237, so the answer is stable.

# 2. Largest gap between short- and long-term rental prices

```{r}
# entire homes with an observed price, nightly stays only, in areas with bond data
gaps <- joined %>%
  filter(room_type == "Entire home/apt",
         price_was_imputed == FALSE,
         long_stay == FALSE,
         has_bond_data == TRUE) %>%
  mutate(long_term_per_night = median_rent_weekly / 7,
         gap = price - long_term_per_night)

nrow(gaps)
n_distinct(gaps$id)
median(gaps$gap)
```

```{r, fig.height = 6}
# gap for every Airbnb, grouped by ward (outliers hidden so the boxes are readable)
par(mar = c(5, 11, 2, 1))
boxplot(gap ~ reorder(neighbourhood, gap, median), data = gaps,
        horizontal = TRUE, las = 1, outline = FALSE, ylab = "",
        xlab = "Airbnb price minus long-term rent ($ per night)")
```

```{r}
# median gap per ward
gaps %>%
  group_by(neighbourhood) %>%
  summarise(listings = n_distinct(id), median_gap = median(gap)) %>%
  arrange(desc(median_gap))

# median gap per location id, areas with at least 20 listings
gaps %>%
  group_by(area_code) %>%
  summarise(ward = first(neighbourhood),
            listings = n_distinct(id),
            airbnb_price = median(price),
            long_term_per_night = round(median(long_term_per_night)),
            median_gap = median(gap)) %>%
  filter(listings >= 20) %>%
  arrange(desc(median_gap)) %>%
  head(5)

# Banks Peninsula has no bond data at all
joined %>%
  filter(neighbourhood == "Banks Peninsula Ward") %>%
  count(has_bond_data)
```

## Result: Heathcote Ward, location 332700

| A whole home in 332700 | Per night |
|---|---|
| Airbnb | $344 |
| Long-term rent | $89 |
| **Gap** | **$255** |

- Christchurch median gap: $144 a night.
- By ward: Halswell is highest ($186) but has only 52 listings. Heathcote is next
  ($164) with 214 listings, and holds the widest area, so we name Heathcote.
- Compares entire homes only, using real prices and nightly stays. An area needs
  20+ listings to be ranked.
- **Caveat:** Banks Peninsula has no bond data (3,548 listing-months). It is full of
  holiday homes, so the real largest gap may be hidden there.

# 3. Number of Airbnbs vs rental properties

```{r}
# latest quarter (Apr-Jun 2026): Airbnb listings vs active bonds in each area
counts <- joined %>%
  filter(quarter == as.Date("2026-04-01"), has_bond_data == TRUE) %>%
  group_by(area_code) %>%
  summarise(ward = first(neighbourhood),
            airbnbs = n_distinct(id),
            rentals = first(bond_active)) %>%
  mutate(airbnbs_per_100_rentals = round(100 * airbnbs / rentals))

sum(counts$airbnbs)
sum(counts$rentals)

counts %>%
  arrange(desc(airbnbs_per_100_rentals)) %>%
  head(5)
```

```{r, fig.height = 6}
# the same comparison per ward
ward_counts <- counts %>%
  group_by(ward) %>%
  summarise(airbnbs = sum(airbnbs), rentals = sum(rentals)) %>%
  mutate(airbnbs_per_100_rentals = round(100 * airbnbs / rentals)) %>%
  arrange(airbnbs_per_100_rentals)

ward_counts

par(mar = c(5, 11, 2, 1))
barplot(ward_counts$airbnbs_per_100_rentals, names.arg = ward_counts$ward,
        horiz = TRUE, las = 1,
        xlab = "Airbnbs per 100 long-term rentals (Apr-Jun 2026)")
```

## Result: 8 Airbnbs per 100 rentals

April to June 2026:

| | Airbnbs | Rentals | Per 100 rentals |
|---|---|---|---|
| All areas with bond data | 2,840 | 36,219 | **8** |
| Central Ward | 1,347 | 7,761 | 17 |
| Christchurch Central (326600) | 146 | 42 | 348 |

- Airbnbs cluster in the centre. Every ward outside Central has 2 to 9 per 100.
- Only two areas have more Airbnbs than rentals: 326600 and 327100 (293 vs 276).
- Counts distinct listings in the quarter against active bonds.
- Banks Peninsula is missing, because it has no bond data.
- **Beds:** cannot be compared. The listings file has no bedrooms column.
