---
title: "Airbnb vs Long-Term Rentals in Christchurch"
subtitle: "Deliverable 5 questions, October 2025 - June 2026 (revised in Deliverable 6)"
output: html_document
---

Answers the three Deliverable 5 questions from `Christchurch_with_bonds.csv`.
Knit `join_listings_bonds.Rmd` first, because it makes that file.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# install once if needed: install.packages(c("dplyr", "readr"))
library(dplyr)
library(readr)
```

```{r settings}
# values used more than once below, so they are set in one place
central_area <- "326600"      # SA2 area code of Christchurch Central
big_host     <- "482671075"   # host with 7 listings at one Central address
min_listings <- 20            # an area needs this many listings to be ranked in question 2
count_month  <- "June 2026"   # the month used to count Airbnbs in question 3
days_per_week <- 7            # bond rents are weekly, Airbnb prices are per night
```

```{r read}
# made by join_listings_bonds.Rmd, so knit that first
joined <- read_csv("Christchurch_with_bonds.csv",
                   col_types = cols(id = col_character(),
                                    host_id = col_character(),
                                    area_code = col_character(),
                                    .default = col_guess()))

# stop if a listing appears twice in the same month
if (anyDuplicated(joined[, c("id", "month_year")]) > 0) {
  stop("The joined file has a listing twice in the same month")
}
```

**How the medians are counted (changed in Deliverable 6).** The joined file has one row
per listing per month. In Deliverable 5 we took medians over rows, so a listing with a
price in six months counted six times and a listing seen once counted once. Now every
listing counts once: we first take each listing's own median price, then the median of
those.

# 1. Median Airbnb price in Christchurch Central (326600)

```{r central}
central <- joined %>% filter(area_code == central_area)

# observed prices only: Dec-Feb have no prices, and prices over $2,000 are errors
observed <- central %>% filter(price_was_imputed == FALSE)

nrow(observed)             # listing-months with a real price
n_distinct(observed$id)    # listings

# how many listings have a real price in 1, 2, ... 6 months
table(table(observed$id))
```

```{r central-median}
# one price per listing: its median over the months it has a real price
central_listings <- observed %>%
  group_by(id) %>%
  summarise(host_id = first(host_id),
            price = median(price))

central_median <- median(central_listings$price)
central_median

# the Deliverable 5 way, every listing-month counted once, for comparison
old_central_median <- median(observed$price)
old_central_median
```

## Does one host skew the median?

Host `r big_host` has 7 listings at one Central address. In Deliverable 5 each of them
counted up to six times, once per month, and the price cleaning had copied one filled
price to all 7 (fixed in `clean_christchurch_panel.Rmd`). We check the median with that
host, and with the biggest hosts, left out.

```{r hosts}
# the hosts with the most Central listings
host_counts <- central_listings %>% count(host_id, sort = TRUE)
head(host_counts, 5)

# share of Central listings run by the three biggest hosts
top3_hosts <- head(host_counts$host_id, 3)
top3_share <- round(100 * mean(central_listings$host_id %in% top3_hosts))
top3_share

# median without the host with 7 listings at one address
median_without_big_host <- median(central_listings$price[central_listings$host_id != big_host])
median_without_big_host

# median without the three biggest hosts
median_without_top3 <- median(central_listings$price[!central_listings$host_id %in% top3_hosts])
median_without_top3
```

```{r central-checks}
# check: include the filled prices (Dec-Feb and the gaps), still one price per listing
central_filled <- central %>%
  group_by(id) %>%
  summarise(price = median(price_imputed))

median(central_filled$price)

# for comparison: all of Christchurch, one price per listing
city_listings <- joined %>%
  filter(price_was_imputed == FALSE) %>%
  group_by(id) %>%
  summarise(price = median(price))

city_median <- median(city_listings$price)
city_median
```

## Result: $`r central_median` per night

| Median price per night | |
|---|---|
| Christchurch Central (`r central_area`) | **$`r central_median`** |
| All of Christchurch | $`r city_median` |
| Central costs more by | $`r central_median - city_median` (`r round(100 * (central_median - city_median) / city_median)`%) |

- Based on `r nrow(observed)` real prices from `r nrow(central_listings)` listings, each
  listing counted once. Counting every listing-month instead gives
  $`r old_central_median`, the Deliverable 5 answer.
- Leaving out host `r big_host` and its 7 listings at one address gives
  $`r median_without_big_host`, so that host does not skew the median once each listing
  counts once.
- The three biggest hosts run `r top3_share`% of Central listings and price below the
  rest. Without them the median would be $`r round(median_without_top3)`. We keep them, because
  every one of their listings is a real place a guest can book.
- Left out: December to February (no prices) and prices over $2,000.

# 2. Largest gap between short- and long-term rental prices

```{r gaps}
# entire homes with an observed price, nightly stays only, in areas with bond data
gaps <- joined %>%
  filter(room_type == "Entire home/apt",
         price_was_imputed == FALSE,
         long_stay == FALSE,
         has_bond_data == TRUE) %>%
  mutate(long_term_per_night = median_rent_weekly / days_per_week,
         gap = price - long_term_per_night)

# one gap per listing and area, its median over the months it has a real price
# (a few listings move area between months, so they get one row per area)
gap_listings <- gaps %>%
  group_by(id, area_code) %>%
  summarise(ward = first(neighbourhood),
            price = median(price),
            long_term_per_night = median(long_term_per_night),
            gap = median(gap)) %>%
  ungroup()

n_distinct(gap_listings$id)       # listings

city_gap <- median(gap_listings$gap)
city_gap
```

```{r gap-plot, fig.height = 6}
# gap for every listing, grouped by ward (outliers hidden so the boxes are readable)
par(mar = c(5, 11, 2, 1))
boxplot(gap ~ reorder(ward, gap, median), data = gap_listings,
        horizontal = TRUE, las = 1, outline = FALSE, ylab = "",
        xlab = "Airbnb price minus long-term rent ($ per night)")
```

```{r gap-tables}
# median gap per ward
ward_gaps <- gap_listings %>%
  group_by(ward) %>%
  summarise(listings = n_distinct(id), median_gap = round(median(gap))) %>%
  arrange(desc(median_gap))

ward_gaps

# median gap per area code, for areas with enough listings
area_gaps <- gap_listings %>%
  group_by(area_code) %>%
  summarise(ward = first(ward),
            listings = n_distinct(id),
            airbnb_price = round(median(price)),
            long_term_per_night = round(median(long_term_per_night)),
            median_gap = round(median(gap))) %>%
  filter(listings >= min_listings) %>%
  arrange(desc(median_gap))

head(area_gaps, 5)

# the two wards named in the result
halswell  <- ward_gaps %>% filter(ward == "Halswell Ward")
heathcote <- ward_gaps %>% filter(ward == "Heathcote Ward")

top_area <- area_gaps[1, ]

# Banks Peninsula has no bond data at all, so it cannot be ranked
joined %>%
  filter(neighbourhood == "Banks Peninsula Ward") %>%
  count(has_bond_data)

banks_rows <- sum(joined$neighbourhood == "Banks Peninsula Ward")
```

## Result: `r top_area$ward`, location `r top_area$area_code`

| A whole home in `r top_area$area_code` | Per night |
|---|---|
| Airbnb | $`r top_area$airbnb_price` |
| Long-term rent | $`r top_area$long_term_per_night` |
| **Gap** | **$`r top_area$median_gap`** |

- Christchurch median gap: $`r round(city_gap)` a night.
- By ward: Halswell is highest ($`r halswell$median_gap`) but has only
  `r halswell$listings` listings. Heathcote is next ($`r heathcote$median_gap`) with
  `r heathcote$listings` listings, and holds the area with the widest gap, so we name
  Heathcote.
- Compares entire homes only, using real prices and nightly stays, with each listing
  counted once. An area needs `r min_listings`+ listings to be ranked.
- The gap is the median of each listing's own gap, so it can differ by a dollar from
  the Airbnb price minus the rent, which are medians taken separately.
- **Caveat:** Banks Peninsula has no bond data (`r format(banks_rows, big.mark = ",")`
  listing-months). It is full of holiday homes, so the real largest gap may be hidden
  there.

# 3. Number of Airbnbs vs rental properties

**Changed in Deliverable 6.** Active bonds count the rentals running at one point in
time. In every area they are far higher than the bonds lodged in the quarter, so they are
not new rentals. In Deliverable 5 we counted every Airbnb seen at any time in April, May
or June. That counts a listing that closed in April and a new one that opened in June as
two Airbnbs, even though they were never listed at the same time. We now count the
Airbnbs listed in one month, June 2026, the last month of the quarter.

```{r counts}
# Airbnbs listed in the count month vs active bonds, in each area with bond data
counts <- joined %>%
  filter(month_year == count_month, has_bond_data == TRUE) %>%
  group_by(area_code) %>%
  summarise(ward = first(neighbourhood),
            airbnbs = n_distinct(id),
            rentals = first(bond_active)) %>%
  mutate(airbnbs_per_100_rentals = round(100 * airbnbs / rentals))

total_airbnbs <- sum(counts$airbnbs)
total_rentals <- sum(counts$rentals)
total_per_100 <- round(100 * total_airbnbs / total_rentals)

total_airbnbs
total_rentals
total_per_100

counts %>%
  arrange(desc(airbnbs_per_100_rentals)) %>%
  head(5)
```

```{r ward-counts, fig.height = 6}
# the same comparison per ward
ward_counts <- counts %>%
  group_by(ward) %>%
  summarise(airbnbs = sum(airbnbs), rentals = sum(rentals)) %>%
  mutate(airbnbs_per_100_rentals = round(100 * airbnbs / rentals)) %>%
  arrange(airbnbs_per_100_rentals)

ward_counts

par(mar = c(5, 11, 2, 1))
barplot(ward_counts$airbnbs_per_100_rentals, names.arg = ward_counts$ward,
        horiz = TRUE, las = 1,
        xlab = paste("Airbnbs per 100 long-term rentals,", count_month))

# rows used in the result table
central_ward <- ward_counts %>% filter(ward == "Central Ward")
central_area_counts <- counts %>% filter(area_code == central_area)
```

## Result: `r total_per_100` Airbnbs per 100 rentals

`r count_month`:

| | Airbnbs | Rentals | Per 100 rentals |
|---|---|---|---|
| All areas with bond data | `r format(total_airbnbs, big.mark = ",")` | `r format(total_rentals, big.mark = ",")` | **`r total_per_100`** |
| Central Ward | `r format(central_ward$airbnbs, big.mark = ",")` | `r format(central_ward$rentals, big.mark = ",")` | `r central_ward$airbnbs_per_100_rentals` |
| Christchurch Central (`r central_area`) | `r central_area_counts$airbnbs` | `r central_area_counts$rentals` | `r central_area_counts$airbnbs_per_100_rentals` |

- Airbnbs cluster in the centre. Every ward outside Central has fewer than 10 per 100.
- Only `r sum(counts$airbnbs > counts$rentals)` areas have more Airbnbs than rentals (see
  the top five areas above).
- Counts distinct listings in `r count_month` against active bonds for the April to
  June 2026 quarter. The Deliverable 5 count over the whole quarter gave 8 per 100.
- Banks Peninsula is missing, because it has no bond data.
- **Beds:** cannot be compared. The listings file has no bedrooms column.
