---
title: "Bond Listing Cleaning"
subtitle: "Tenancy Services detailed quarterly report, filtered to Oct 2025 - Jun 2026"
date: "2026-09-13"
output: html_document
---

Cleans the Tenancy Services bond report, `Detailed-Quarterly-Tenancy-Q1-2020-Q3-2026.csv`,
and writes `Bond_Data_Quarterly_(cleaned).csv`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(lubridate)
```

# 1. Read the Data

```{r settings}
# file paths are relative: "." is the repo folder, so this works on every computer
data_dir <- "."

in_file <- file.path(data_dir, 
                     "Detailed-Quarterly-Tenancy-Q1-2020-Q3-2026.csv")

listings_file <- file.path(data_dir, 
                           "Christchurch Oct2025 to Jun2026 (cleaned).csv") 
#Reads the listings file only to get its date range for filtering months from quarterly

# same name as the file join_listings_bonds.Rmd reads
out_file <- file.path(data_dir, "Bond_Data_Quarterly_(cleaned).csv")

# the Location Id the bond report uses for its all-of-New-Zealand total
national_total_id <- "-99"

if (!file.exists(in_file)) stop("Input file not found: ", in_file)
if (!file.exists(listings_file)) stop("Input file not found: ", listings_file)
```

```{r read}
# Location Id is an area code, a label, so it is read as text.
# Read as a number, area 200000 was saved as "2e+05".
bonds <- read_csv(
  in_file,
  na = c("", "NA", "NULL"),
  col_types = cols(`Location Id` = col_character(), .default = col_guess())
)
dim(bonds)
```

# 2. Match the timeframe to the listings panel

Bond data is reported quarterly. The listings panel is monthly. Instead of
hardcoding which quarters to keep, we read the cleaned listings output and
matched them against its actual date range. So this filter is still correct if 
the listings panel is edited.

```{r timeframe}

airbnb <- read_csv(listings_file, show_col_types = FALSE)

listings_start <- min(airbnb$month_date)
listings_end   <- max(airbnb$month_date)

c(listings_start, listings_end)
```

```{r matching_quarters}

bonds <- bonds %>% mutate(TimeFrame = as.Date(TimeFrame))

quarter_lookup <- bonds %>%
  distinct(TimeFrame) %>%
  mutate(quarter_end = TimeFrame %m+% months(3) - days(1)) # We calculate quarter_end so we can check if the quarter's full 3-month span overlaps the listings' date range, not just whether its start date falls inside it. Calculates the last day of each bond quarter. 

target_quarters <- quarter_lookup %>%
  filter(TimeFrame <= listings_end, quarter_end >= listings_start) %>%
  pull(TimeFrame) # Extracts the TimeFrame column as a plain list of values (not a table), so it can be used with %in% in the next step.

target_quarters

bonds <- bonds %>% filter(TimeFrame %in% target_quarters)
dim(bonds)
```

Three quarters overlap the listings window: 
2025-10-01 (Oct-Dec 2025),2026-01-01 (Jan-Mar 2026) and 2026-04-01 (Apr-Jun 2026). 
There is no 2026-07-01 quarter yet, since bond data is published a couple of months
behind the current date and does not yet reach June 2026.

# 3. Column selection

Unlike the listings data, every column here is kept. 
`Total Bonds`,`Active Bonds` and `Closed Bonds` 
are the stock-side variables needed. `Median Rent`and the other rent statistics 
are the price-side variables. Dropping any of them now would remove information 
the merge specifically needs.

# 4. Location Id

```{r location_diagnosis}
sum(is.na(bonds$`Location Id`)) 
sum(bonds$`Location Id` == national_total_id, na.rm = TRUE)
```

Missing Location Id. These rows are also blank on every rent
statistic, so they carry no usable information beyond the row existing at
all, and there is no defensible value to impute a location with.

```{r location_drop}
bonds <- bonds %>% filter(!is.na(`Location Id`), `Location Id` != national_total_id)
dim(bonds)
```

When Location Id is -99 it represents an "All of New Zealand"
rollup, not a real SA2 area. Keeping it would mix a national total into
what should be a location-level dataset, and it cannot be matched to
anything in a location-based merge, so it is dropped rather than treated
as a location like any of the other values.

# 5. Number Of Beds

## 5.1 What the missing bed counts are

```{r beds-diagnose}
sum(is.na(bonds$`Number Of Beds`))

bonds %>%
  group_by(`Dwelling Type`) %>%
  summarise(rows = n(), missing = sum(is.na(`Number Of Beds`)))

# does the same area, quarter and dwelling type also have its own "ALL" total row?
all_rows <- bonds %>%
  filter(`Number Of Beds` == "ALL") %>%
  select(TimeFrame, `Location Id`, `Dwelling Type`)

bonds %>%
  filter(is.na(`Number Of Beds`)) %>%
  semi_join(all_rows, by = c("TimeFrame", "Location Id", "Dwelling Type")) %>%
  nrow()
```

Almost every row with a missing bed count sits next to its own `"ALL"` total row and has
its own bond counts. So a missing bed count is not a gap in a normal row. It is a
separate group: bonds where the number of bedrooms was not recorded.

## 5.2 Our approach (changed in Deliverable 6)

In Deliverable 4 we filled these with kNN. kNN gave them a bed count the area already
had, so the file ended up with two rows for the same area, quarter, dwelling type and bed
count (779 of them), and 44 areas with two `"ALL"` total rows. The Rmd said there were no
duplicates, but its own check printed 779.

Now we label them `"Not recorded"`, which is what they are. Nothing is guessed, and every
row is unique again.

```{r beds-label}
bonds <- bonds %>%
  mutate(`Number Of Beds` = ifelse(is.na(`Number Of Beds`), "Not recorded",
                                   `Number Of Beds`))

table(bonds$`Number Of Beds`)
```

# 6. Validation

```{r checks}
key_columns <- c("TimeFrame", "Location Id", "Dwelling Type", "Number Of Beds")

if (anyDuplicated(bonds[, key_columns]) > 0) {
  stop("Two rows share the same quarter, area, dwelling type and bed count")
}
if (anyNA(bonds$`Location Id`)) stop("Some rows have no Location Id")
if (!all(bonds$TimeFrame %in% target_quarters)) stop("A row is outside the target quarters")

range(bonds$TimeFrame)
length(unique(bonds$`Location Id`))
colSums(is.na(bonds))
```

If the knit reaches this point, there are no duplicate rows on the natural key, every
row has an area, and every row is in one of the three target quarters. The rent columns
can still be blank, because Tenancy Services leaves them out when an area has too few
bonds.

# 7. Known limitations

1. "Location Id" is an SA2 area code, not a name. The listings get their SA2 area
   codes from `get_area_codes.py`, and the two are joined in `join_listings_bonds.Rmd`.

2. Bond data is quarterly while listings are monthly, so any merge will
   necessarily compare a single quarterly figure against up to three
   monthly listings figures.
  
  
# 8. Save

```{r save}
bonds <- bonds %>% arrange(TimeFrame, `Location Id`)
write_csv(bonds, out_file)

out_file
nrow(bonds)
```
   
   
