---
title: "Bond Listing Cleaning"
subtitle: "Tenancy Services detailed quarterly report, filtered to Oct 2025 - Jun 2026"
date: "2026-09-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(readr)
library(VIM)
library(lubridate)
```

# 1. Read the Data

```{r paths}

data_dir <- "/Users/nalikadesai/Desktop/DATA201"

in_file       <- file.path(data_dir, 
                           "Detailed-Quarterly-Tenancy-Q1-2020-Q3-2026.csv")

listings_file <- file.path(data_dir, 
                           "Christchurch Oct2025 to Jun2026 (cleaned).csv") 
#Reads the listings file only to get its date range for filtering months from quarterly

out_file <- file.path(data_dir, "Bond Data Quarterly (cleaned).csv")

file.exists(in_file)
file.exists(listings_file)
```

```{r read}
bonds <- read_csv(in_file, show_col_types = FALSE)
dim(bonds)
```

# 2. Match the timeframe to the listings panel

Bond data is reported quarterly. The listings panel is monthly. Instead of
hardcoding which quarters to keep, we read the cleaned listings output and
matched them against its actual date range. So this filter is still correct if 
the listings panel is edited.

```{r timeframe}

airbnb <- read_csv(listings_file, show_col_types = FALSE)

listings_start <- min(airbnb$month_date)
listings_end   <- max(airbnb$month_date)

c(listings_start, listings_end)
```

```{r matching_quarters}

bonds <- bonds %>% mutate(TimeFrame = as.Date(TimeFrame))

quarter_lookup <- bonds %>%
  distinct(TimeFrame) %>%
  mutate(quarter_end = TimeFrame %m+% months(3) - days(1)) # Adds 3 months to the quarter's start date, then subtracts 1 day, to get that quarter's last day (e.g. 2025-10-01 -> 2025-12-31).

target_quarters <- quarter_lookup %>%
  filter(TimeFrame <= listings_end, quarter_end >= listings_start) %>%
  pull(TimeFrame) # Extracts the TimeFrame column as a plain list of values (not a table), so it can be used with %in% in the next step.

target_quarters

bonds <- bonds %>% filter(TimeFrame %in% target_quarters)
dim(bonds)
```

Three quarters overlap the listings window: 
2025-10-01 (Oct-Dec 2025),2026-01-01 (Jan-Mar 2026) and 2026-04-01 (Apr-Jun 2026). 
There is no 2026-07-01 quarter yet, since bond data is published a couple of months
behind the current date and does not yet reach June 2026.

# 3. Column selection

Unlike the listings data, every column here is kept. 
`Total Bonds`,`Active Bonds` and `Closed Bonds` 
are the stock-side variables needed. `Median Rent`and the other rent statistics 
are the price-side variables. Dropping any of them now would remove information 
the merge specifically needs.

# 4. Location Id

```{r location_diagnosis}
sum(is.na(bonds$`Location Id`)) 
sum(bonds$`Location Id` == -99, na.rm = TRUE)
```

Missing Location Id. These rows are also blank on every rent
statistic, so they carry no usable information beyond the row existing at
all, and there is no defensible value to impute a location with.

```{r location_drop}
bonds <- bonds %>% filter(!is.na(`Location Id`), `Location Id` != -99)
dim(bonds)
```

When Location Id == -99 it represents an "All of New Zealand"
rollup, not a real SA2 area. Keeping it would mix a national total into
what should be a location-level dataset, and it cannot be matched to
anything in a location-based merge, so it is dropped rather than treated
as a location like any of the other values.

# 5. Types

```{r types}
bonds <- bonds %>%
  mutate(
    `Dwelling Type` = as.factor(`Dwelling Type`),
    `Number Of Beds` = as.factor(`Number Of Beds`),
    `Location Id` = as.factor(`Location Id`)
  )
```

# 6. Number Of Beds

## 6.1 What the missingness looks like

```{r beds-diagnose}
sum(is.na(bonds$`Number Of Beds`))

bonds %>%
  group_by(`Dwelling Type`) %>%
  summarise(rows = n(), missing = sum(is.na(`Number Of Beds`)),
            pct_missing = round(mean(is.na(`Number Of Beds`)) * 100, 1)) # For each Dwelling Type, counts total rows, how many are missing Number Of Beds, and what percent that is — to check if missingness is concentrated in one type or spread out evenly.
```

This is distinct from the existing "ALL" factor level (a legitimate
aggregate-across-beds row). These are individual rows where the specific
bed-count breakdown was not reported, spread thinly across dwelling
types rather than concentrated in one category or quarter.


## 6.2 Our approach

We discussed this as a group and decided to use kNN imputation for the missing prices.
It is one method among several, and we may test other approaches later to compare the
results.

As with price in the listings data, we impute using kNN: for each row
missing "Number Of Beds", we find the most similar rows on dwelling type,
location, bond counts, and rent statistics, and take their bed count. A
house with similar rent and bond activity in the same area is a reasonable
proxy for bedroom count.

```{r beds-knn}
set.seed(2026)

bonds <- kNN(
  bonds,
  variable = "Number Of Beds",
  dist_var = c("Dwelling Type", "Location Id", "Total Bonds", "Active Bonds",
               "Closed Bonds", "Median Rent", "Geometric Mean Rent",
               "Upper Quartile Rent", "Lower Quartile Rent"),
  k = 5,
  imp_var = TRUE
)

bonds <- bonds %>% rename(beds_was_imputed = `Number Of Beds_imp`)

sum(bonds$beds_was_imputed)
sum(is.na(bonds$`Number Of Beds`))
```

The original data had no separate "unimputed" copy of this column to
preserve (unlike price, which stayed at 0% missing in some months and
scattered elsewhere) -- every row now has a value, and `beds_was_imputed`
marks which ones were filled in.

# 7. Validation

```{r checks}
sum(duplicated(bonds[, c("TimeFrame", "Location Id", "Dwelling Type", "Number Of Beds")]))
range(bonds$TimeFrame)
length(unique(bonds$`Location Id`))
colSums(is.na(bonds))
```

No duplicate rows on the natural key, "TimeFrame" falls entirely within
the three target quarters, and no missing values remain in any column.


# 8. Known limitations

1. "Location Id" is an SA2 area code, not a geographic coordinate or a
   name. There is no location-name lookup in this file, so it cannot yet
   be filtered to Christchurch-only rows or joined to the listings data
   by name. A separate SA2-to-district lookup (e.g. from Stats NZ) will
   be needed before next week's merge.
   
2. The kNN distance calculation treats `Location Id` as a categorical
   match (same area vs. different area), not as spatial distance --
   different SA2 codes are not numerically "close" to each other in any
   meaningful sense, so cross-location neighbour matching leans more on
   dwelling type and bond/rent figures than true geographic proximity.
   
3. Bond data is quarterly while listings are monthly, so any merge will
   necessarily compare a single quarterly figure against up to three
   monthly listings figures.
  
  
# 9. Save

```{r save}
bonds <- bonds %>% arrange(TimeFrame, `Location Id`)
write_csv(bonds, out_file)

out_file
nrow(bonds)
```
   
   
