---
title: "Cleaning and Imputing the Christchurch Airbnb Panel"
subtitle: "Inside Airbnb listings, October 2025 - June 2026"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Cleans the combined Airbnb panel, `Christchurch Oct2025 to Jun2026 (combined).csv`, and
writes `Christchurch Oct2025 to Jun2026 (cleaned).csv`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# install once if needed: install.packages(c("dplyr", "tidyr", "readr", "VIM"))
library(dplyr)
library(tidyr)
library(readr)
library(VIM)
```

# 1. Read the data

```{r settings}
# file paths are relative: "." is the repo folder, so this works on every computer
# (in VS Code, open the repo folder first: File > Open Folder)
data_dir <- "."
in_file  <- file.path(data_dir, "Christchurch Oct2025 to Jun2026 (combined).csv")
out_file <- file.path(data_dir, "Christchurch Oct2025 to Jun2026 (cleaned).csv")

# cleaning rules, set once here instead of as numbers inside the code
price_max        <- 2000   # a nightly price above this is an error
knn_k            <- 10     # how many similar listings kNN uses to fill a price
long_stay_nights <- 30     # a minimum stay this long makes a listing long-stay
random_seed      <- 2026   # makes kNN give the same result every run

if (!file.exists(in_file)) stop("Input file not found: ", in_file)
```

```{r read}
airbnb <- read_csv(
  in_file,
  col_types = cols(id = col_character(), host_id = col_character(), .default = col_guess())
)

dim(airbnb)
```

**Why character IDs.** The largest `id` is 1,709,885,808,209,520,396. A double holds
about 15 significant digits, so reading these as numbers silently corrupts the last few.
They are labels, never arithmetic, so character is the correct type. This is the same
issue that produced the type mismatch in the KNIME workflow.

# 2. Drop the empty column

```{r drop}
sum(is.na(airbnb$license))

airbnb <- airbnb %>% select(-license)
```

`license` is `NA` in all 28,795 rows. A column with zero observed values cannot
inform anything and only adds noise to the missing-value summary.

`neighbourhood_group` is kept even though it is the constant `"Christchurch City"`,

# 3. Month as an ordered factor and a real date

```{r month}
month_levels <- c("October 2025", "November 2025", "December 2025",
                  "January 2026", "February 2026", "March 2026",
                  "April 2026", "May 2026", "June 2026")

airbnb <- airbnb %>%
  mutate(month_date = case_when(
    month_year == "October 2025"  ~ as.Date("2025-10-01"),
    month_year == "November 2025" ~ as.Date("2025-11-01"),
    month_year == "December 2025" ~ as.Date("2025-12-01"),
    month_year == "January 2026"  ~ as.Date("2026-01-01"),
    month_year == "February 2026" ~ as.Date("2026-02-01"),
    month_year == "March 2026"    ~ as.Date("2026-03-01"),
    month_year == "April 2026"    ~ as.Date("2026-04-01"),
    month_year == "May 2026"      ~ as.Date("2026-05-01"),
    month_year == "June 2026"     ~ as.Date("2026-06-01")
  )) %>%
  mutate(month_year = factor(month_year, levels = month_levels, ordered = TRUE))

table(airbnb$month_year)
```

As plain text the months sort alphabetically, so every grouped summary and
every x-axis came out April, August, December, February. The ordered factor fixes
plotting and grouping; `month_date` allows arithmetic and correct time-series joins.

# 4. Dates

```{r dates}
airbnb <- airbnb %>% mutate(last_review = as.Date(last_review))
range(airbnb$last_review, na.rm = TRUE)
```


# 5. Reviews for listings that have none

```{r reviews}
sum(airbnb$number_of_reviews == 0 & is.na(airbnb$reviews_per_month))

airbnb <- airbnb %>%
  mutate(reviews_per_month = ifelse(number_of_reviews == 0, 0, reviews_per_month))
```

**Why 0 rather than imputation.** All 2,627 missing values sit on listings with
`number_of_reviews == 0`. This is a structural blank, not a collection failure: a
listing with no reviews genuinely receives 0 reviews per month. Imputing a market
average would invent review activity that never happened.

**Why `last_review` stays `NA`.** There is no correct date to insert. A never-reviewed
listing has no last review, and any placeholder would corrupt days-since-review.

# 6. Host name

```{r host}
airbnb <- airbnb %>%
  arrange(host_id, month_date) %>%
  group_by(host_id) %>%
  fill(host_name, .direction = "downup") %>%
  ungroup() %>%
  mutate(host_name = ifelse(is.na(host_name), "Unknown", host_name))

sum(is.na(airbnb$host_name))
```

**Why.** One row was missing a host name. That `host_id` appears 9 times overall and is
named in the other 8, so the value is recovered with certainty rather than guessed.

# 7. Minimum nights

```{r minnights}
airbnb <- airbnb %>%
  arrange(id, month_date) %>%
  group_by(id) %>%
  mutate(minimum_nights_filled = minimum_nights) %>%
  fill(minimum_nights_filled, .direction = "downup") %>%
  ungroup() %>%
  mutate(minimum_nights_was_filled = is.na(minimum_nights) & !is.na(minimum_nights_filled))

sum(airbnb$minimum_nights_was_filled)
```

**Why carry rather than impute.** `minimum_nights` is a booking rule the host sets, not
a market outcome. It changes rarely, and all 37 gaps had a value from the same listing
in an adjacent month. The original column is kept alongside.

# 8. Price

Price is the only column needing a real imputation decision.

## 8.1 What the missingness looks like

```{r price-diagnose}
airbnb %>%
  group_by(month_year) %>%
  summarise(rows = n(), price_observed = sum(!is.na(price)),
            pct_missing = round(mean(is.na(price)) * 100, 1))
```

Two distinct problems:

* **December 2025, January 2026 and February 2026 have no prices at all.** Not a sample
  of missing values, the entire column is blank in those three snapshots.
* **The other six months are 5-8% missing**, scattered across listings.

## 8.2 The scattered gaps are not missing at random

```{r mnar}
airbnb %>%
  filter(!month_year %in% c("December 2025", "January 2026", "February 2026")) %>%
  group_by(price_missing = is.na(price)) %>%
  summarise(rows = n(),
            pct_zero_availability = round(mean(availability_365 == 0) * 100, 1))
```

Among rows with a price, 0.5% have zero availability. Among rows missing a price, 62%
do. Inside Airbnb derives price from the booking calendar, so a listing with no
available nights has no price to report.

## 8.3 Our approach (updated in Deliverable 6)

In Deliverable 4 we filled every missing price with kNN, using the most similar listings
on room type, location, minimum nights, how many listings the host runs, and month.

**What went wrong.** kNN cannot tell apart listings that have the same host and the same
address, because they match on every one of those columns. Host 482671075 runs 7
listings at one address in Christchurch Central, from 2-bedroom flats to 4-bedroom
townhouses. kNN gave all 7 the same price, $213 in December and January and $315 in
February. The 4-bedroom townhouse charges $450 to $741 in the months with real prices.
In total, 564 filled prices were one value copied across a host's listings like this.

**The fix: fill from the listing itself first.** Most missing prices are in December to
February, and the same listing has real prices in the other months. The listing's own
median price is a better guess than the price of a neighbour. So a missing price is now
filled in this order:

1. the real price, if there is one
2. the listing's own median price from the months it does have (10,176 rows)
3. kNN, with the same settings as before, only for listings with no real price in any
   month (518 rows)

```{r price-knn}
set.seed(random_seed)

# prices over price_max are errors, so treat them as missing
sum(airbnb$price > price_max, na.rm = TRUE)

airbnb <- airbnb %>%
  mutate(price_real = ifelse(price > price_max, NA, price))

# kNN guess for every missing price, same settings as Deliverable 4
airbnb <- airbnb %>%
  mutate(price_knn = price_real)

airbnb <- kNN(airbnb,
              variable = "price_knn",
              dist_var = c("room_type", "latitude", "longitude",
                           "minimum_nights_filled",
                           "calculated_host_listings_count", "month_year"),
              k = knn_k, imp_var = FALSE)

# each listing's own median price over the months it has a real price
# (NA if the listing never has a real price)
airbnb <- airbnb %>%
  group_by(id) %>%
  mutate(own_median_price = median(price_real, na.rm = TRUE)) %>%
  ungroup()

# record where each price comes from, then pick the value in that order
airbnb <- airbnb %>%
  mutate(price_source = case_when(
    !is.na(price_real)       ~ "observed",
    !is.na(own_median_price) ~ "own median",
    TRUE                     ~ "kNN"
  )) %>%
  mutate(price_imputed = case_when(
    price_source == "observed"   ~ price_real,
    price_source == "own median" ~ own_median_price,
    price_source == "kNN"        ~ price_knn
  )) %>%
  mutate(price_was_imputed = price_source != "observed")

table(airbnb$price_source)
if (anyNA(airbnb$price_imputed)) stop("Some rows still have no price")
```

The original `price` column is kept unchanged, `price_imputed` holds the filled version,
`price_was_imputed` marks which rows differ, and the new `price_source` column says how
each price was filled.

```{r price-check-host}
# the 7 listings at the shared address now keep their own prices in December
airbnb %>%
  filter(host_id == "482671075", month_year == "December 2025",
         latitude == -43.535568, longitude == 172.6318853) %>%
  select(name, price_knn, price_imputed, price_source)
```

**Sanity check one row by hand.** Take the 4-bedroom townhouse from the example above.
Its real prices are $614, $615, $741, $490, $450 and $450. In order they are
450, 450, 490, 614, 615, 741, so its median is (490 + 614) / 2 = **$552**. Its filled
December price should be exactly that.

```{r price-check-one-row}
townhouse_december <- airbnb %>%
  filter(id == "967760129147377073", month_year == "December 2025")

townhouse_december$price_imputed

if (townhouse_december$price_imputed != 552) {
  stop("The townhouse's filled December price is not its own median of $552")
}
```

```{r price-check-copies}
# filled prices that are one value copied across a host's listings at the same
# address and month: first with kNN alone, then with the new method
copies <- airbnb %>%
  filter(price_was_imputed) %>%
  group_by(host_id, latitude, longitude, month_year) %>%
  mutate(listings = n(),
         knn_prices = n_distinct(price_knn),
         new_prices = n_distinct(price_imputed)) %>%
  ungroup()

sum(copies$listings > 1 & copies$knn_prices == 1)   # kNN alone
sum(copies$listings > 1 & copies$new_prices == 1)   # now

# the helper columns are not needed in the saved file
airbnb <- airbnb %>% select(-price_real, -price_knn, -own_median_price)
```

```{r price-check}
airbnb %>%
  group_by(month_year) %>%
  summarise(rows = n(),
            observed = sum(!is.na(price)),
            imputed  = sum(price_was_imputed),
            median_price = round(median(price_imputed), 0))
```

# 9. Panel structure and remaining flags

```{r panel}
airbnb <- airbnb %>%
  group_by(id) %>%
  mutate(months_present = n_distinct(month_year)) %>%
  ungroup() %>%
  mutate(in_all_9_months = months_present == length(month_levels),
         long_stay = minimum_nights_filled >= long_stay_nights,
         name = trimws(name))

table(airbnb$months_present)
```

**Why this matters.** Only 2,338 of 4,117 listings appear in all nine snapshots, so an
unbalanced panel mixes real price movement with listings entering and leaving. Any
month-over-month comparison should either use `in_all_9_months` or state that it does not.

`long_stay` marks the 129 rows requiring 30 or more nights. These are a different market
to nightly rental and are worth excluding from tourist-price analysis.

# 10. Validation

```{r checks}
if (anyDuplicated(airbnb[, c("id", "month_year")]) > 0) {
  stop("A listing appears twice in the same month")
}
range(airbnb$latitude)
range(airbnb$longitude)
colSums(is.na(airbnb))
```

Coordinates fall between -43.87 and -43.42 latitude and 172.43 to 173.10 longitude,
which is Christchurch City including Banks Peninsula. No stray points.

# 11. Known limitations

1. December 2025 to February 2026 contain no observed prices, so those values are the
   least reliable in the dataset. December to February is the summer peak, and a
   listing's own median comes from the other months, so filled summer prices are
   probably a little too low.
2. The 518 rows filled by kNN belong to listings that never show a real price, so they
   still rely on neighbours and can still share one value across a host's listings.
3. Imputation uncertainty is not carried through, so standard errors computed on
   `price_imputed` will be slightly too small.

# 12. Save

```{r save}
airbnb <- airbnb %>% arrange(month_date, id)
write_csv(airbnb, out_file)

out_file
nrow(airbnb)
```
