---
title: "Cleaning and Imputing the Christchurch Airbnb Panel"
subtitle: "Inside Airbnb listings, October 2025 - June 2026"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# install once if needed: install.packages(c("dplyr", "tidyr", "readr", "VIM"))
library(dplyr)
library(tidyr)
library(readr)
library(VIM)
```

# 1. Read the data

```{r paths}
# "." means the folder this Rmd is in (the repo folder), so it works on every computer.
# Knitting, and running chunks in RStudio, both use that folder automatically.
# Running chunks in VS Code: open the repo folder first (File > Open Folder).
data_dir <- "."

in_file  <- file.path(data_dir, "Christchurch Oct2025 to Jun2026 (combined).csv")
out_file <- file.path(data_dir, "Christchurch Oct2025 to Jun2026 (cleaned).csv")

file.exists(in_file)
```

```{r read}
airbnb <- read_csv(
  in_file,
  col_types = cols(id = col_character(), host_id = col_character(), .default = col_guess())
)

dim(airbnb)
```

**Why character IDs.** The largest `id` is 1,709,885,808,209,520,396. A double holds
about 15 significant digits, so reading these as numbers silently corrupts the last few.
They are labels, never arithmetic, so character is the correct type. This is the same
issue that produced the type mismatch in the KNIME workflow.

# 2. Drop the empty column

```{r drop}
sum(is.na(airbnb$license))

airbnb <- airbnb %>% select(-license)
```

`license` is `NA` in all 28,795 rows. A column with zero observed values cannot
inform anything and only adds noise to the missing-value summary.

`neighbourhood_group` is kept even though it is the constant `"Christchurch City"`,

# 3. Month as an ordered factor and a real date

```{r month}
month_levels <- c("October 2025", "November 2025", "December 2025",
                  "January 2026", "February 2026", "March 2026",
                  "April 2026", "May 2026", "June 2026")

airbnb <- airbnb %>%
  mutate(month_date = case_when(
    month_year == "October 2025"  ~ as.Date("2025-10-01"),
    month_year == "November 2025" ~ as.Date("2025-11-01"),
    month_year == "December 2025" ~ as.Date("2025-12-01"),
    month_year == "January 2026"  ~ as.Date("2026-01-01"),
    month_year == "February 2026" ~ as.Date("2026-02-01"),
    month_year == "March 2026"    ~ as.Date("2026-03-01"),
    month_year == "April 2026"    ~ as.Date("2026-04-01"),
    month_year == "May 2026"      ~ as.Date("2026-05-01"),
    month_year == "June 2026"     ~ as.Date("2026-06-01")
  )) %>%
  mutate(month_year = factor(month_year, levels = month_levels, ordered = TRUE))

table(airbnb$month_year)
```

As plain text the months sort alphabetically, so every grouped summary and
every x-axis came out April, August, December, February. The ordered factor fixes
plotting and grouping; `month_date` allows arithmetic and correct time-series joins.

# 4. Dates

```{r dates}
airbnb <- airbnb %>% mutate(last_review = as.Date(last_review))
range(airbnb$last_review, na.rm = TRUE)
```


# 5. Reviews for listings that have none

```{r reviews}
sum(airbnb$number_of_reviews == 0 & is.na(airbnb$reviews_per_month))

airbnb <- airbnb %>%
  mutate(reviews_per_month = ifelse(number_of_reviews == 0, 0, reviews_per_month))
```

**Why 0 rather than imputation.** All 2,627 missing values sit on listings with
`number_of_reviews == 0`. This is a structural blank, not a collection failure: a
listing with no reviews genuinely receives 0 reviews per month. Imputing a market
average would invent review activity that never happened.

**Why `last_review` stays `NA`.** There is no correct date to insert. A never-reviewed
listing has no last review, and any placeholder would corrupt days-since-review.

# 6. Host name

```{r host}
airbnb <- airbnb %>%
  arrange(host_id, month_date) %>%
  group_by(host_id) %>%
  fill(host_name, .direction = "downup") %>%
  ungroup() %>%
  mutate(host_name = ifelse(is.na(host_name), "Unknown", host_name))

sum(is.na(airbnb$host_name))
```

**Why.** One row was missing a host name. That `host_id` appears 9 times overall and is
named in the other 8, so the value is recovered with certainty rather than guessed.

# 7. Minimum nights

```{r minnights}
airbnb <- airbnb %>%
  arrange(id, month_date) %>%
  group_by(id) %>%
  mutate(minimum_nights_filled = minimum_nights) %>%
  fill(minimum_nights_filled, .direction = "downup") %>%
  ungroup() %>%
  mutate(minimum_nights_was_filled = is.na(minimum_nights) & !is.na(minimum_nights_filled))

sum(airbnb$minimum_nights_was_filled)
```

**Why carry rather than impute.** `minimum_nights` is a booking rule the host sets, not
a market outcome. It changes rarely, and all 37 gaps had a value from the same listing
in an adjacent month. The original column is kept alongside.

# 8. Price

Price is the only column needing a real imputation decision.

## 8.1 What the missingness looks like

```{r price-diagnose}
airbnb %>%
  group_by(month_year) %>%
  summarise(rows = n(), price_observed = sum(!is.na(price)),
            pct_missing = round(mean(is.na(price)) * 100, 1))
```

Two distinct problems:

* **December 2025, January 2026 and February 2026 have no prices at all.** Not a sample
  of missing values, the entire column is blank in those three snapshots.
* **The other six months are 5-8% missing**, scattered across listings.

## 8.2 The scattered gaps are not missing at random

```{r mnar}
airbnb %>%
  filter(!month_year %in% c("December 2025", "January 2026", "February 2026")) %>%
  group_by(price_missing = is.na(price)) %>%
  summarise(rows = n(),
            pct_zero_availability = round(mean(availability_365 == 0) * 100, 1))
```

Among rows with a price, 0.5% have zero availability. Among rows missing a price, 62%
do. Inside Airbnb derives price from the booking calendar, so a listing with no
available nights has no price to report.

## 8.3 Our approach

We discussed this as a group and decided to use kNN imputation for the missing prices.
It is one method among several, and we may test other approaches later to compare the
results.

kNN fills each missing price using the most similar listings, judged on room type,
location, minimum nights and how many listings the host runs. Month is included so
same-month listings are preferred as neighbours. Because December to February contain no
prices at all, values for those months are drawn from listings in other months.

```{r price-knn}
set.seed(2026)

airbnb <- airbnb %>%
  mutate(price_imputed = ifelse(price > 2000, NA, price))   # 27 impossible values

airbnb <- kNN(airbnb,
              variable = "price_imputed",
              dist_var = c("room_type", "latitude", "longitude",
                           "minimum_nights_filled",
                           "calculated_host_listings_count", "month_year"),
              k = 10, imp_var = FALSE)

airbnb <- airbnb %>%
  mutate(price_was_imputed = is.na(price) | price > 2000)

sum(airbnb$price_was_imputed)
sum(is.na(airbnb$price_imputed))
```

The original `price` column is kept unchanged, `price_imputed` holds the filled version,
and `price_was_imputed` marks which rows differ.

```{r price-check}
airbnb %>%
  group_by(month_year) %>%
  summarise(rows = n(),
            observed = sum(!is.na(price)),
            imputed  = sum(price_was_imputed),
            median_price = round(median(price_imputed), 0))
```

# 9. Panel structure and remaining flags

```{r panel}
airbnb <- airbnb %>%
  group_by(id) %>%
  mutate(months_present = n_distinct(month_year)) %>%
  ungroup() %>%
  mutate(in_all_9_months = months_present == 9,
         long_stay = minimum_nights_filled >= 30,
         name = trimws(name))

table(airbnb$months_present)
```

**Why this matters.** Only 2,338 of 4,117 listings appear in all nine snapshots, so an
unbalanced panel mixes real price movement with listings entering and leaving. Any
month-over-month comparison should either use `in_all_9_months` or state that it does not.

`long_stay` marks the 129 rows requiring 30 or more nights. These are a different market
to nightly rental and are worth excluding from tourist-price analysis.

# 10. Validation

```{r checks}
sum(duplicated(airbnb[, c("id", "month_year")]))   # expect 0
range(airbnb$latitude)
range(airbnb$longitude)
colSums(is.na(airbnb))
```

Coordinates fall between -43.87 and -43.42 latitude and 172.43 to 173.10 longitude,
which is Christchurch City including Banks Peninsula. No stray points.

# 11. Known limitations

1. December 2025 to February 2026 contain no observed prices, so those values are the
   least reliable in the dataset.
2. kNN was chosen as a first approach. Other imputation methods have not been tested yet
   and may give different results.
3. Imputation uncertainty is not carried through, so standard errors computed on
   `price_imputed` will be slightly too small.

# 12. Save

```{r save}
airbnb <- airbnb %>% arrange(month_date, id)
write_csv(airbnb, out_file)

out_file
nrow(airbnb)
```
