---
title: "Joining the Airbnb Listings to the Bond Data"
subtitle: "Christchurch, October 2025 - June 2026, joined on area code and quarter"
output: html_document
---

Joins the cleaned listings to the cleaned bond totals by area code and quarter, and
writes `Christchurch_with_bonds.csv`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# install once if needed: install.packages(c("dplyr", "readr"))
library(dplyr)
library(readr)
```

Run the files in this order: `clean_christchurch_panel.Rmd`, `bond_listing_clean.Rmd`,
then this file, then `airbnb_vs_rentals.Rmd`. `get_area_codes.py` only needs to run again
if the listing coordinates change.

# 1. Read the files

```{r read}
# the cleaned listings, made by clean_christchurch_panel.Rmd
listings <- read_csv("Christchurch Oct2025 to Jun2026 (cleaned).csv",
                     col_types = cols(id = col_character(),
                                      host_id = col_character(),
                                      .default = col_guess()))

# the area code of every listing, made by get_area_codes.py
# area codes are labels, so they are read as text and match exactly
area_codes <- read_csv("Christchurch_with_area_codes.csv",
                       col_types = cols(id = col_character(),
                                        host_id = col_character(),
                                        area_code = col_character(),
                                        .default = col_guess()))

# the bond file writes missing values as the word NULL
bonds <- read_csv("Bond_Data_Quarterly_(cleaned).csv",
                  na = c("", "NA", "NULL"),
                  col_types = cols(`Location Id` = col_character(),
                                   .default = col_guess()))

dim(listings)
dim(area_codes)
dim(bonds)
```

# 2. Add the area codes

`get_area_codes.py` asks an online map service for the area code of every coordinate.
It needs an API key, so it is not run every time. Its output is a full copy of the
cleaned listings plus an `area_code` column. That copy goes out of date whenever the
cleaning changes, which happened in Deliverable 6 when the price filling was fixed.

So we take only the `area_code` column and attach it to the latest cleaned listings by
listing `id` and month. Area codes depend only on coordinates, and the cleaning does not
change any coordinates.

```{r area-codes}
area_codes <- area_codes %>% select(id, month_year, area_code)

if (anyDuplicated(area_codes[, c("id", "month_year")]) > 0) {
  stop("A listing has two area codes in the same month, so the join would add rows")
}

listings <- listings %>% left_join(area_codes, by = c("id", "month_year"))

if (anyNA(listings$area_code)) stop("Some listings have no area code")
```

# 3. Give every listing its quarter

Listings are monthly and bonds are quarterly, so each month gets the first day of its
quarter, which is what the bond `TimeFrame` holds.

```{r quarter}
listings <- listings %>%
  mutate(quarter = case_when(
    month_year %in% c("October 2025", "November 2025", "December 2025") ~ as.Date("2025-10-01"),
    month_year %in% c("January 2026", "February 2026", "March 2026")    ~ as.Date("2026-01-01"),
    month_year %in% c("April 2026", "May 2026", "June 2026")            ~ as.Date("2026-04-01")
  ))

if (anyNA(listings$quarter)) stop("Some months were not given a quarter")
```

# 4. Keep one bond row per area and quarter

The bond file has a row for every dwelling type and bed count, so an area has about
five rows per quarter. We keep the area total, which is the row where both are
`"ALL"`.

```{r bond-totals}
bond_totals <- bonds %>%
  filter(`Dwelling Type` == "ALL",
         `Number Of Beds` == "ALL") %>%
  select(area_code                  = `Location Id`,
         quarter                    = TimeFrame,
         bond_total                 = `Total Bonds`,
         bond_active                = `Active Bonds`,
         bond_closed                = `Closed Bonds`,
         median_rent_weekly         = `Median Rent`,
         geo_mean_rent_weekly       = `Geometric Mean Rent`,
         upper_quartile_rent_weekly = `Upper Quartile Rent`,
         lower_quartile_rent_weekly = `Lower Quartile Rent`,
         log_sd_rent_weekly         = `Log Std Dev Weekly Rent`)

nrow(bond_totals)

if (anyDuplicated(bond_totals[, c("area_code", "quarter")]) > 0) {
  stop("An area has two total rows in the same quarter, so the join would add rows")
}
```

# 5. Join

A left join keeps every Airbnb row. An inner join would drop listings and undercount
Airbnbs per area, and a full join would add bond areas from all of New Zealand.

```{r join}
joined <- listings %>%
  left_join(bond_totals, by = c("area_code", "quarter")) %>%
  mutate(has_bond_data = !is.na(bond_total))

if (nrow(joined) != nrow(listings)) stop("The join changed the number of rows")

table(joined$has_bond_data)
```

# 6. Why some rows have no bond data

```{r no-bond}
# every area code in the raw bond file, 2020 to 2026, not just our three quarters
raw_bonds <- read_csv("Detailed-Quarterly-Tenancy-Q1-2020-Q3-2026.csv",
                      col_types = cols(.default = col_character()))
bond_areas_ever <- unique(raw_bonds$`Location Id`)

# every area and quarter pair that has at least one bond row
bond_area_quarters <- unique(paste(bonds$`Location Id`, bonds$TimeFrame))

no_bond_reasons <- joined %>%
  filter(has_bond_data == FALSE) %>%
  mutate(reason = case_when(
    !area_code %in% bond_areas_ever                     ~ "1. area code never appears in the bond data",
    !paste(area_code, quarter) %in% bond_area_quarters ~ "2. area has no bond rows that quarter",
    TRUE                                                ~ "3. area has bond rows that quarter, but no total row"
  )) %>%
  count(reason)

no_bond_reasons
```

1. The area code never appears in the bond data, 2020 to 2026. These are mostly areas
   split after the bond data's boundaries were drawn (listings have 332501, bonds have
   332500).
2. The area is in the bond data, but had too few bonds to be published that quarter.
3. The area has bond rows that quarter, but not the `"ALL"` total row we join to.

In Deliverable 5 these counts were typed in by hand (4,862, 1,465 and 522). Computing them
here shows reason 3 was wrong: most of those 522 rows have no bond rows at all that
quarter, so they belong under reason 2.

# 7. Sanity check one row by hand

Take one Christchurch Central (326600) listing in April 2026. April is in the quarter
that starts on 2026-04-01, so the listing should carry the bond total for area 326600 in
that quarter. We look that row up straight from the raw bond report, not from our
cleaned file, and compare the two.

```{r one-row}
one_listing <- joined %>%
  filter(area_code == "326600", month_year == "April 2026") %>%
  head(1)

one_listing %>% select(id, month_year, quarter, bond_active, median_rent_weekly)

raw_total <- raw_bonds %>%
  filter(`Location Id` == "326600", TimeFrame == "2026-04-01",
         `Dwelling Type` == "ALL", `Number Of Beds` == "ALL")

raw_total %>% select(TimeFrame, `Location Id`, `Active Bonds`, `Median Rent`)

if (one_listing$median_rent_weekly != as.numeric(raw_total$`Median Rent`) ||
    one_listing$bond_active != as.numeric(raw_total$`Active Bonds`)) {
  stop("The joined bond values do not match the raw bond report for this row")
}
```

# 8. Save

```{r save}
write_csv(joined, "Christchurch_with_bonds.csv")

dim(joined)
```

# Result

`r format(sum(joined$has_bond_data), big.mark = ",")` listing rows
(`r round(100 * mean(joined$has_bond_data))`%) matched a bond total. The other
`r format(sum(!joined$has_bond_data), big.mark = ",")` are kept with empty bond columns,
for the three reasons counted in section 6. Most are areas split after the bond data's
boundaries were drawn.

Notes for using the joined file:

- Rents are **weekly**, prices are **per night**. Divide rents by 7 before comparing.
- Only rows where `has_bond_data` is `TRUE` can compare prices with rents.
- A listing appears once a month, so count distinct `id`s, not rows.
- 37 listings change area code between months, because Airbnb shifts map locations
  slightly for privacy.
