---
title: "Joining the Airbnb Listings to the Bond Data"
subtitle: "Christchurch, October 2025 - June 2026, joined on area code and quarter"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# install once if needed: install.packages(c("dplyr", "readr"))
library(dplyr)
library(readr)
```

# 1. Read the two files

```{r}
# area codes are labels, so both sides are read as text and match exactly
# the bond file writes missing values as the word NULL
listings <- read_csv("Christchurch_with_area_codes.csv",
                     col_types = cols(id = col_character(),
                                      host_id = col_character(),
                                      area_code = col_character(),
                                      .default = col_guess()))

bonds <- read_csv("Bond_Data_Quarterly_(cleaned).csv",
                  na = c("", "NA", "NULL"),
                  col_types = cols(`Location Id` = col_character(),
                                   .default = col_guess()))

dim(listings)
dim(bonds)
```

# 2. Give every listing its quarter

Listings are monthly and bonds are quarterly, so each month gets the first day of its
quarter, which is what the bond `TimeFrame` holds.

```{r}
listings <- listings %>%
  mutate(quarter = case_when(
    month_year %in% c("October 2025", "November 2025", "December 2025") ~ as.Date("2025-10-01"),
    month_year %in% c("January 2026", "February 2026", "March 2026")    ~ as.Date("2026-01-01"),
    month_year %in% c("April 2026", "May 2026", "June 2026")            ~ as.Date("2026-04-01")
  ))

sum(is.na(listings$quarter))   # expect 0
```

# 3. Keep one bond row per area and quarter

The bond file has a row for every dwelling type and bed count, so an area has about
five rows per quarter. We keep the area total, which is the row where both are
`"ALL"`. The kNN step in `bond_listing_clean.Rmd` filled 20 missing bed counts with
`"ALL"`, which makes those rows look like totals, so imputed rows are dropped too.

```{r}
bond_totals <- bonds %>%
  filter(`Dwelling Type` == "ALL",
         `Number Of Beds` == "ALL",
         beds_was_imputed == FALSE) %>%
  select(area_code                  = `Location Id`,
         quarter                    = TimeFrame,
         bond_total                 = `Total Bonds`,
         bond_active                = `Active Bonds`,
         bond_closed                = `Closed Bonds`,
         median_rent_weekly         = `Median Rent`,
         geo_mean_rent_weekly       = `Geometric Mean Rent`,
         upper_quartile_rent_weekly = `Upper Quartile Rent`,
         lower_quartile_rent_weekly = `Lower Quartile Rent`,
         log_sd_rent_weekly         = `Log Std Dev Weekly Rent`)

nrow(bond_totals)
sum(duplicated(bond_totals[, c("area_code", "quarter")]))   # expect 0
```

# 4. Join

A left join keeps every Airbnb row. An inner join would drop listings and undercount
Airbnbs per area, and a full join would add bond areas from all of New Zealand.

```{r}
joined <- listings %>%
  left_join(bond_totals, by = c("area_code", "quarter")) %>%
  mutate(has_bond_data = !is.na(bond_total))

nrow(joined) == nrow(listings)   # expect TRUE, the join adds no rows

table(joined$has_bond_data)
```

# 5. Save

```{r}
write_csv(joined, "Christchurch_with_bonds.csv")

dim(joined)
```

# Result

21,946 listing rows (76%) matched a bond total. The other 6,849 are kept with empty
bond columns, for three reasons:

| Reason | Rows |
|---|---|
| Area code is not in the bond data, mostly areas split since its boundaries (listings have 332501, bonds have 332500) | 4,862 |
| Area is in the bond data, but is too quiet to be published in these quarters | 1,465 |
| Area has bond rows that quarter, but no total row | 522 |

Notes for using the joined file:

- Rents are **weekly**, prices are **per night**. Divide rents by 7 before comparing.
- Only rows where `has_bond_data` is `TRUE` can compare prices with rents.
- A listing appears once a month, so count distinct `id`s per quarter, not rows.
- 37 listings change area code between months, because Airbnb shifts map locations
  slightly for privacy.
