---
title: "Joining the Airbnb Listings to the Bond Data"
subtitle: "Christchurch, October 2025 - June 2026, joined on area code and quarter"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# install once if needed: install.packages(c("dplyr", "readr"))
library(dplyr)
library(readr)
```

Run the files in this order: `clean_christchurch_panel.Rmd`, `bond_listing_clean.Rmd`,
then this file, then `airbnb_vs_rentals.Rmd`. `get_area_codes.py` only needs to run again
if the listing coordinates change.

# 1. Read the files

```{r read}
# the cleaned listings, made by clean_christchurch_panel.Rmd
listings <- read_csv("Christchurch Oct2025 to Jun2026 (cleaned).csv",
                     col_types = cols(id = col_character(),
                                      host_id = col_character(),
                                      .default = col_guess()))

# the area code of every listing, made by get_area_codes.py
# area codes are labels, so they are read as text and match exactly
area_codes <- read_csv("Christchurch_with_area_codes.csv",
                       col_types = cols(id = col_character(),
                                        host_id = col_character(),
                                        area_code = col_character(),
                                        .default = col_guess()))

# the bond file writes missing values as the word NULL
bonds <- read_csv("Bond_Data_Quarterly_(cleaned).csv",
                  na = c("", "NA", "NULL"),
                  col_types = cols(`Location Id` = col_character(),
                                   .default = col_guess()))

dim(listings)
dim(area_codes)
dim(bonds)
```

# 2. Add the area codes

`get_area_codes.py` asks an online map service for the area code of every coordinate.
It needs an API key, so it is not run every time. Its output is a full copy of the
cleaned listings plus an `area_code` column. That copy goes out of date whenever the
cleaning changes, which happened in Deliverable 6 when the price filling was fixed.

So we take only the `area_code` column and attach it to the latest cleaned listings by
listing `id` and month. Area codes depend only on coordinates, and the cleaning does not
change any coordinates.

```{r area-codes}
area_codes <- area_codes %>% select(id, month_year, area_code)

# one area code per listing per month, so the join cannot add rows
stopifnot(sum(duplicated(area_codes[, c("id", "month_year")])) == 0)

listings <- listings %>% left_join(area_codes, by = c("id", "month_year"))

stopifnot(sum(is.na(listings$area_code)) == 0)   # every listing got an area code
```

# 3. Give every listing its quarter

Listings are monthly and bonds are quarterly, so each month gets the first day of its
quarter, which is what the bond `TimeFrame` holds.

```{r quarter}
listings <- listings %>%
  mutate(quarter = case_when(
    month_year %in% c("October 2025", "November 2025", "December 2025") ~ as.Date("2025-10-01"),
    month_year %in% c("January 2026", "February 2026", "March 2026")    ~ as.Date("2026-01-01"),
    month_year %in% c("April 2026", "May 2026", "June 2026")            ~ as.Date("2026-04-01")
  ))

stopifnot(sum(is.na(listings$quarter)) == 0)   # every month got a quarter
```

# 4. Keep one bond row per area and quarter

The bond file has a row for every dwelling type and bed count, so an area has about
five rows per quarter. We keep the area total, which is the row where both are
`"ALL"`. The kNN step in `bond_listing_clean.Rmd` filled some missing bed counts with
`"ALL"`, which makes those rows look like totals, so imputed rows are dropped too.

```{r bond-totals}
# kNN rows that look like area totals, dropped below
sum(bonds$beds_was_imputed & bonds$`Dwelling Type` == "ALL" & bonds$`Number Of Beds` == "ALL")

bond_totals <- bonds %>%
  filter(`Dwelling Type` == "ALL",
         `Number Of Beds` == "ALL",
         beds_was_imputed == FALSE) %>%
  select(area_code                  = `Location Id`,
         quarter                    = TimeFrame,
         bond_total                 = `Total Bonds`,
         bond_active                = `Active Bonds`,
         bond_closed                = `Closed Bonds`,
         median_rent_weekly         = `Median Rent`,
         geo_mean_rent_weekly       = `Geometric Mean Rent`,
         upper_quartile_rent_weekly = `Upper Quartile Rent`,
         lower_quartile_rent_weekly = `Lower Quartile Rent`,
         log_sd_rent_weekly         = `Log Std Dev Weekly Rent`)

nrow(bond_totals)

# one total per area and quarter, so the join cannot add rows
stopifnot(sum(duplicated(bond_totals[, c("area_code", "quarter")])) == 0)
```

# 5. Join

A left join keeps every Airbnb row. An inner join would drop listings and undercount
Airbnbs per area, and a full join would add bond areas from all of New Zealand.

```{r join}
joined <- listings %>%
  left_join(bond_totals, by = c("area_code", "quarter")) %>%
  mutate(has_bond_data = !is.na(bond_total))

stopifnot(nrow(joined) == nrow(listings))   # the join adds no rows

table(joined$has_bond_data)
```

# 6. Why some rows have no bond data

```{r no-bond}
# every area code in the raw bond file, 2020 to 2026, not just our three quarters
raw_bonds <- read_csv("Detailed-Quarterly-Tenancy-Q1-2020-Q3-2026.csv",
                      col_types = cols(.default = col_character()))
bond_areas_ever <- unique(raw_bonds$`Location Id`)

# every area and quarter pair that has at least one bond row
bond_area_quarters <- unique(paste(bonds$`Location Id`, bonds$TimeFrame))

no_bond_reasons <- joined %>%
  filter(has_bond_data == FALSE) %>%
  mutate(reason = case_when(
    !area_code %in% bond_areas_ever                     ~ "1. area code never appears in the bond data",
    !paste(area_code, quarter) %in% bond_area_quarters ~ "2. area has no bond rows that quarter",
    TRUE                                                ~ "3. area has bond rows that quarter, but no total row"
  )) %>%
  count(reason)

no_bond_reasons
```

1. The area code never appears in the bond data, 2020 to 2026. These are mostly areas
   split after the bond data's boundaries were drawn (listings have 332501, bonds have
   332500).
2. The area is in the bond data, but had too few bonds to be published that quarter.
3. The area has bond rows that quarter, but not the `"ALL"` total row we join to.

In Deliverable 5 these counts were typed in by hand (4,862, 1,465 and 522). Computing them
here shows reason 3 was wrong: most of those 522 rows have no bond rows at all that
quarter, so they belong under reason 2.

# 7. Save

```{r save}
write_csv(joined, "Christchurch_with_bonds.csv")

dim(joined)
```

# Result

`r format(sum(joined$has_bond_data), big.mark = ",")` listing rows
(`r round(100 * mean(joined$has_bond_data))`%) matched a bond total. The other
`r format(sum(!joined$has_bond_data), big.mark = ",")` are kept with empty bond columns,
for the three reasons counted in section 6. Most are areas split after the bond data's
boundaries were drawn.

Notes for using the joined file:

- Rents are **weekly**, prices are **per night**. Divide rents by 7 before comparing.
- Only rows where `has_bond_data` is `TRUE` can compare prices with rents.
- A listing appears once a month, so count distinct `id`s, not rows.
- 37 listings change area code between months, because Airbnb shifts map locations
  slightly for privacy.
