---
title: "Joining the Airbnb Listings to the Bond Data"
subtitle: "Christchurch, October 2025 - June 2026, joined on area code and quarter"
output: html_document
---

Reads `Christchurch_with_area_codes.csv` and `Bond_Data_Quarterly_(cleaned).csv`, joins
them on area code and quarter, and saves `Christchurch_with_bonds.csv`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# install once if needed: install.packages(c("dplyr", "readr"))
library(dplyr)
library(readr)
```

# 1. Read the two files

```{r settings}
listings_file  <- "Christchurch_with_area_codes.csv"
bonds_file     <- "Bond_Data_Quarterly_(cleaned).csv"
raw_bonds_file <- "Detailed-Quarterly-Tenancy-Q1-2020-Q3-2026.csv"   # only used for the sanity check
out_file       <- "Christchurch_with_bonds.csv"
```

```{r}
# area codes are labels, so both sides are read as text and match exactly
# the bond file writes missing values as the word NULL
listings <- read_csv(listings_file,
                     col_types = cols(id = col_character(),
                                      host_id = col_character(),
                                      area_code = col_character(),
                                      .default = col_guess()))

bonds <- read_csv(bonds_file,
                  na = c("", "NA", "NULL"),
                  col_types = cols(`Location Id` = col_character(),
                                   .default = col_guess()))

dim(listings)
dim(bonds)
```

# 2. Give every listing its quarter

Listings are monthly and bonds are quarterly, so each month gets the first day of its
quarter, which is what the bond `TimeFrame` holds.

```{r}
listings <- listings %>%
  mutate(quarter = case_when(
    month_year %in% c("October 2025", "November 2025", "December 2025") ~ as.Date("2025-10-01"),
    month_year %in% c("January 2026", "February 2026", "March 2026")    ~ as.Date("2026-01-01"),
    month_year %in% c("April 2026", "May 2026", "June 2026")            ~ as.Date("2026-04-01")
  ))

if (anyNA(listings$quarter)) stop("Some months were not given a quarter")
```

# 3. Keep one bond row per area and quarter

The bond file has a row for every dwelling type and bed count, so an area has about
five rows per quarter. We keep the area total, which is the row where both are
`"ALL"`. The kNN step in `bond_listing_clean.Rmd` filled 20 missing bed counts with
`"ALL"`, which makes those rows look like totals, so imputed rows are dropped too.

```{r}
bond_totals <- bonds %>%
  filter(`Dwelling Type` == "ALL",
         `Number Of Beds` == "ALL",
         beds_was_imputed == FALSE) %>%
  select(area_code                  = `Location Id`,
         quarter                    = TimeFrame,
         bond_total                 = `Total Bonds`,
         bond_active                = `Active Bonds`,
         bond_closed                = `Closed Bonds`,
         median_rent_weekly         = `Median Rent`,
         geo_mean_rent_weekly       = `Geometric Mean Rent`,
         upper_quartile_rent_weekly = `Upper Quartile Rent`,
         lower_quartile_rent_weekly = `Lower Quartile Rent`,
         log_sd_rent_weekly         = `Log Std Dev Weekly Rent`)

nrow(bond_totals)
if (anyDuplicated(bond_totals[, c("area_code", "quarter")]) > 0) {
  stop("An area has two bond totals in the same quarter, so the join would add rows")
}
```

# 4. Join

A left join keeps every Airbnb row. An inner join would drop listings and undercount
Airbnbs per area, and a full join would add bond areas from all of New Zealand.

```{r}
joined <- listings %>%
  left_join(bond_totals, by = c("area_code", "quarter")) %>%
  mutate(has_bond_data = !is.na(bond_total))

if (nrow(joined) != nrow(listings)) stop("The join changed the number of rows")

table(joined$has_bond_data)
```

# 5. Sanity check one row by hand

We pick one Christchurch Central (326600) listing in April 2026. April is in the quarter
that starts on 2026-04-01, so the join should give it the area total for 326600 in that
quarter. We look that row up ourselves in the raw bond report, not in our cleaned file,
and check the two agree.

```{r sanity-check}
one_listing <- joined %>%
  filter(area_code == "326600", month_year == "April 2026") %>%
  head(1)

one_listing %>% select(id, month_year, quarter, bond_active, median_rent_weekly)

raw_bonds <- read_csv(raw_bonds_file, col_types = cols(.default = col_character()))

raw_row <- raw_bonds %>%
  filter(`Location Id` == "326600", TimeFrame == "2026-04-01",
         `Dwelling Type` == "ALL", `Number Of Beds` == "ALL")

raw_row %>% select(TimeFrame, `Location Id`, `Active Bonds`, `Median Rent`)

if (one_listing$bond_active != as.numeric(raw_row$`Active Bonds`) ||
    one_listing$median_rent_weekly != as.numeric(raw_row$`Median Rent`)) {
  stop("The joined bond numbers do not match the raw bond report")
}
```

Both show `r one_listing$bond_active` active bonds and a weekly rent of
$`r one_listing$median_rent_weekly`, so the join matched the right area and the right quarter.

# 6. Save

```{r}
write_csv(joined, out_file)

dim(joined)
```

# Result

21,946 listing rows (76%) matched a bond total. The other 6,849 are kept with empty
bond columns, for three reasons:

| Reason | Rows |
|---|---|
| Area code is not in the bond data, mostly areas split since its boundaries (listings have 332501, bonds have 332500) | 4,862 |
| Area is in the bond data, but is too quiet to be published in these quarters | 1,465 |
| Area has bond rows that quarter, but no total row | 522 |

Notes for using the joined file:

- Rents are **weekly**, prices are **per night**. Divide rents by 7 before comparing.
- Only rows where `has_bond_data` is `TRUE` can compare prices with rents.
- A listing appears once a month, so count distinct `id`s per quarter, not rows.
- 37 listings change area code between months, because Airbnb shifts map locations
  slightly for privacy.
