---
title: "Airbnb vs Long-Term Rentals in Christchurch"
subtitle: "Deliverable 5 questions, every month from October 2025"
output: html_document
---

Reads `output/Christchurch_with_bonds.csv` and answers the three Deliverable 5 questions.
Knit `join_listings_bonds.Rmd` first, because it makes that file (or run `make`, which
runs every step in order).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# this file is in src/, so run its code from the repo folder (one level up)
knitr::opts_knit$set(root.dir = "..")
library(dplyr)
library(readr)
```

```{r settings}
central_area   <- "326600"                # SA2 area code of Christchurch Central
min_listings   <- 20                      # an area needs this many listings to be ranked
latest_quarter <- as.Date("2026-04-01")   # April to June 2026
days_per_week  <- 7                       # rents are weekly, Airbnb prices are per night
```

```{r}
# made by join_listings_bonds.Rmd, so knit that first
joined <- read_csv("output/Christchurch_with_bonds.csv",
                   col_types = cols(id = col_character(),
                                    host_id = col_character(),
                                    area_code = col_character(),
                                    .default = col_guess()))

# Check that the input contains every column this report needs.
required_columns <- c(
  "id",
  "month_year",
  "month_date",
  "area_code",
  "neighbourhood",
  "price",
  "price_imputed",
  "price_was_imputed",
  "room_type",
  "long_stay",
  "has_bond_data",
  "median_rent_weekly",
  "quarter",
  "bond_active"
)

missing_columns <- setdiff(required_columns, names(joined))

if (length(missing_columns) > 0) {
  stop(
    "The input CSV is missing required columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

if (anyDuplicated(joined[, c("id", "month_year")]) > 0) {
  stop("A listing appears twice in the same month")
}
```

# 1. Median Airbnb price in Christchurch Central (326600)

```{r}
central <- joined %>% filter(area_code == central_area)

nrow(central)             # listing-months
n_distinct(central$id)    # listings
```

```{r}
# observed prices only: Dec-Feb have no prices, and prices over $2,000 are errors
observed <- central %>% filter(price_was_imputed == FALSE)

nrow(observed)
median(observed$price)
```

```{r}
# check 1: include the imputed prices
median(central$price_imputed)

# check 2: count each listing once
per_listing <- observed %>%
  group_by(id) %>%
  summarise(price = median(price))

median(per_listing$price)

# for comparison: all of Christchurch
all_christchurch <- median(joined$price[joined$price_was_imputed == FALSE])
all_christchurch

# the numbers for the result below
central_median <- median(observed$price)
difference     <- central_median - all_christchurch
```

## Result: $`r round(central_median)` per night

| Median price per night | |
|---|---|
| Christchurch Central (326600) | **$`r round(central_median)`** |
| All of Christchurch | $`r round(all_christchurch)` |
| Central costs more by | $`r round(difference)` (`r round(100 * difference / all_christchurch)`%) |

- Based on `r nrow(observed)` real prices from `r n_distinct(observed$id)` listings.
- Left out: December to February (no prices) and `r sum(central$price > 2000, na.rm = TRUE)` prices over $2,000.
- Counting other ways gives $`r round(median(central$price_imputed))` and $`r round(median(per_listing$price))`, so the answer is stable.
- `r sum(is.na(joined$area_code))` listing rows in new months have no area code yet, so they are not counted here.

# 2. Largest gap between short- and long-term rental prices

```{r}
# entire homes with an observed price, nightly stays only, in areas with bond data
gaps <- joined %>%
  filter(room_type == "Entire home/apt",
         price_was_imputed == FALSE,
         long_stay == FALSE,
         has_bond_data == TRUE) %>%
  mutate(long_term_per_night = median_rent_weekly / days_per_week,
         gap = price - long_term_per_night)

nrow(gaps)
n_distinct(gaps$id)
median(gaps$gap)
```

```{r, fig.height = 6}
# gap for every Airbnb, grouped by ward (outliers hidden so the boxes are readable)
par(mar = c(5, 11, 2, 1))
boxplot(gap ~ reorder(neighbourhood, gap, median), data = gaps,
        horizontal = TRUE, las = 1, outline = FALSE, ylab = "",
        xlab = "Airbnb price minus long-term rent ($ per night)")
```

```{r gap_rankings}
# Median gap for each ward.
ward_gaps <- gaps %>%
  group_by(neighbourhood) %>%
  summarise(
    listings = n_distinct(id),
    median_gap = median(gap),
    .groups = "drop"
  ) %>%
  arrange(desc(median_gap))

ward_gaps

# Rank areas with enough distinct listings.
area_gaps <- gaps %>%
  group_by(area_code) %>%
  summarise(
    ward = first(neighbourhood),
    listings = n_distinct(id),
    airbnb_price = median(price),
    long_term_per_night = median(long_term_per_night),
    median_gap = median(gap),
    .groups = "drop"
  ) %>%
  filter(listings >= min_listings, is.finite(median_gap)) %>%
  arrange(desc(median_gap), area_code)

head(area_gaps, 5)

if (nrow(area_gaps) == 0) {
  stop("No areas have enough listings and a valid median gap.")
}

top_area <- slice_head(area_gaps, n = 1)
city_median_gap <- median(gaps$gap)
latest_bond_quarter <- max(gaps$quarter)
```

## Result: Area `r top_area$area_code`, associated ward: `r top_area$ward`

| Entire homes in area `r top_area$area_code` | Per night |
|---|---|
| Median Airbnb price | $`r round(top_area$airbnb_price)` |
| Median daily equivalent of long-term rent | $`r round(top_area$long_term_per_night)` |
| **Median listing-level gap** | **$`r round(top_area$median_gap)`** |

- Christchurch's median listing-level gap is $`r round(city_median_gap)` per night.
- The highest-ranked area contains `r top_area$listings` distinct Airbnb listings.
- Areas need at least `r min_listings` distinct listings to enter the ranking.
- The comparison uses entire homes, observed prices, nightly stays and available bond data.
- The latest included bond quarter starts on `r format(latest_bond_quarter, "%d %B %Y")`.
- The median gap is calculated from individual listing-level gaps. It does not necessarily equal the difference between the two displayed medians.

# 3. Number of Airbnbs vs rental properties

```{r rental_counts}
# Select the comparison quarter and areas with bond data.
quarter_data <- joined %>%
  filter(
    quarter == latest_quarter,
    has_bond_data == TRUE
  )

if (nrow(quarter_data) == 0) {
  stop("No rows with bond data are available for the selected quarter.")
}

# Check area codes and rental counts before dividing.
invalid_rentals <- quarter_data %>%
  filter(
    is.na(area_code) |
    is.na(bond_active) |
    !is.finite(bond_active) |
    bond_active <= 0
  )

if (nrow(invalid_rentals) > 0) {
  stop("Missing area codes or invalid rental counts. Check the joined data.")
}

# Each area needs one consistent rental count for this quarter.
rental_consistency <- quarter_data %>%
  group_by(area_code) %>%
  summarise(
    rental_values = n_distinct(bond_active),
    .groups = "drop"
  )

if (any(rental_consistency$rental_values != 1)) {
  stop("An area has conflicting rental counts within the selected quarter.")
}

# Calculate ratios after the checks pass.
counts <- quarter_data %>%
  group_by(area_code) %>%
  summarise(
    ward = first(neighbourhood),
    airbnbs = n_distinct(id),
    rentals = first(bond_active),
    .groups = "drop"
  ) %>%
  mutate(
    airbnbs_per_100_rentals = round(100 * airbnbs / rentals)
  )

sum(counts$airbnbs)
sum(counts$rentals)

counts %>%
  arrange(desc(airbnbs_per_100_rentals)) %>%
  head(5)
```

```{r, fig.height = 6}
# the same comparison per ward
ward_counts <- counts %>%
  group_by(ward) %>%
  summarise(airbnbs = sum(airbnbs), rentals = sum(rentals)) %>%
  mutate(airbnbs_per_100_rentals = round(100 * airbnbs / rentals)) %>%
  arrange(airbnbs_per_100_rentals)

ward_counts

par(mar = c(5, 11, 2, 1))
barplot(ward_counts$airbnbs_per_100_rentals, names.arg = ward_counts$ward,
        horiz = TRUE, las = 1,
        xlab = "Airbnbs per 100 long-term rentals (Apr-Jun 2026)")
```

```{r count_results}
# Use the calculated counts to produce the report text and table.
total_airbnbs <- sum(counts$airbnbs)
total_rentals <- sum(counts$rentals)

if (is.na(total_rentals) || total_rentals <= 0) {
  stop("The total rental count must be present and positive.")
}

overall_ratio <- round(100 * total_airbnbs / total_rentals)

central_ward <- ward_counts %>%
  filter(ward == "Central Ward")

central_location <- counts %>%
  filter(area_code == central_area)

if (nrow(central_ward) != 1 || nrow(central_location) != 1) {
  stop("Expected one Central Ward row and one Christchurch Central row.")
}

quarter_start <- format(latest_quarter, "%B")
quarter_end <- format(
  seq(latest_quarter, by = "month", length.out = 3)[3],
  "%B %Y"
)

areas_above_100 <- sum(counts$airbnbs > counts$rentals)
```

## Result: `r overall_ratio` Airbnbs per 100 rentals

`r quarter_start` to `r quarter_end`:

| | Airbnbs | Rentals | Per 100 rentals |
|---|---|---|---|
| All areas with bond data | `r format(total_airbnbs, big.mark = ",")` | `r format(total_rentals, big.mark = ",")` | **`r overall_ratio`** |
| Central Ward | `r central_ward$airbnbs` | `r central_ward$rentals` | `r central_ward$airbnbs_per_100_rentals` |
| Christchurch Central (`r central_area`) | `r central_location$airbnbs` | `r central_location$rentals` | `r central_location$airbnbs_per_100_rentals` |

- There are `r areas_above_100` areas with more Airbnb listings than active rental bonds.
- Airbnb counts represent distinct listing IDs observed during the quarter.
- The comparison includes only areas with bond data.

# 4. Prices by month

Added in Deliverable 7, so the new months show up in a plot. Questions 2 and 3 need bond
data, which comes out a few months late, so they only change once Tenancy Services
publishes the next quarter.

```{r}
# median real price per month: all of Christchurch and Christchurch Central
by_month <- joined %>%
  group_by(month_date) %>%
  summarise(listings = n_distinct(id),
            all_christchurch = median(price[price_was_imputed == FALSE]),
            central = median(price[price_was_imputed == FALSE & area_code %in% central_area]))

by_month
```

```{r, fig.width = 9, fig.height = 5.5}
# the first and last month in the data, for the title
first_month <- format(min(by_month$month_date), "%B %Y")
last_month  <- format(max(by_month$month_date), "%B %Y")

# December to February have no real prices, so the lines have a gap there
plot(by_month$month_date, by_month$central, type = "b", pch = 17, col = "#eb6834",
     ylim = range(c(by_month$central, by_month$all_christchurch), na.rm = TRUE) + c(-10, 10),
     xaxt = "n", xlab = "", ylab = "Median price ($ per night)",
     main = paste("Median real Airbnb price per night,", first_month, "to", last_month))
lines(by_month$month_date, by_month$all_christchurch, type = "b", pch = 16, col = "#2a78d6")

# a label for every month in the data, so a new month gets its own label
axis(1, at = by_month$month_date, labels = format(by_month$month_date, "%b\n%Y"),
     padj = 0.5, cex.axis = 0.8)

# explain the gap, for the months with no real prices
no_prices <- by_month$month_date[is.na(by_month$all_christchurch)]
if (length(no_prices) > 0) {
  text(mean(no_prices), mean(range(by_month$all_christchurch, na.rm = TRUE)),
       "no real prices\nin these months", col = "grey40", cex = 0.9)
}

# write the prices of the newest month next to its points
last <- nrow(by_month)
text(by_month$month_date[last], by_month$central[last],
     paste0("$", round(by_month$central[last])), pos = 3, col = "#eb6834", cex = 0.9)
text(by_month$month_date[last], by_month$all_christchurch[last],
     paste0("$", round(by_month$all_christchurch[last])), pos = 1, col = "#2a78d6", cex = 0.9)

legend("topleft", legend = c("Christchurch Central (326600)", "All of Christchurch"),
       col = c("#eb6834", "#2a78d6"), pch = c(17, 16), lty = 1, bty = "n")
```

# Interpretation and limitations

Airbnb prices are advertised nightly prices, not confirmed booking revenue.
We divide weekly long-term rent by seven to express both prices on a daily
basis. However, the rental types differ in tenancy conditions, included
services and additional charges. The price gap should not be interpreted
as a difference in profit.

Active rental bonds are used as a proxy for long-term rental properties.
Airbnb counts represent distinct listing IDs observed during the selected
quarter. They do not show how many properties were available on one
specific day. These counts therefore measure different things.

The observed-price analysis excludes imputed prices. Months without
observed prices do not contribute to these estimates. In the pooled
analysis, listings with prices recorded in more months contribute more
observations.

Each comparison includes only records with the area codes and bond data
needed for that question. The results describe the included records.
They do not establish that Airbnb causes changes in long-term rents.
