---
title: "Combining the Monthly Airbnb Files"
subtitle: "Inside Airbnb listings for Christchurch, every month from October 2025"
output: html_document
---

Reads every monthly Inside Airbnb file in `data/listings/`, keeps the Christchurch rows,
and saves them all as one file: `output/Christchurch_combined.csv`.

This does the same job as `Filter CHC.R` and `Combine CHC.R` from Deliverable 3, but with a
loop, so a new month needs no new code. To add a month, save its `listings.csv` in
`data/listings/`, named by its scrape date (for example `2026-08-13.csv`), then run `make`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# this file is in src/, so run its code from the repo folder (one level up)
knitr::opts_knit$set(root.dir = "..")
```

# 1. The monthly files

Each file is one month of New Zealand listings. Its name is the date Inside Airbnb collected
the data (the scrape date), so the files are listed in date order.

```{r files}
files <- list.files("data/listings", pattern = "\\.csv$", full.names = TRUE)

files
```

# 2. Read each month and keep Christchurch

The loop goes through the files one at a time. For each month it keeps the Christchurch
rows, works out the month from the file name, and adds the rows to one table.

```{r combine}
chch <- NULL   # start with nothing, then add one month at a time

for (f in files) {

  month <- read.csv(f,
                    na.strings = c("", "NA"),
                    colClasses = c(id = "character", host_id = "character"))

  month <- subset(month, neighbourhood_group == "Christchurch City")

  # the month comes from the scrape date at the start of the file name
  scrape_date <- as.Date(substr(basename(f), 1, 10), format = "%Y-%m-%d")

  if (is.na(scrape_date)) {
    stop("This file is not named by its scrape date, like 2026-08-13.csv: ", f)
  }

  month$month_year <- format(scrape_date, "%B %Y")
  month$month_date <- as.Date(format(scrape_date, "%Y-%m-01"))

  chch <- rbind(chch, month)
}
```

`id` and `host_id` are read as text, because the ids are too long to store as numbers
without losing digits. `month_date` is the first day of the month, so later files can sort
and group by month correctly.

# 3. Check

```{r check}
# two files from the same month would put every listing in twice
if (length(unique(chch$month_year)) != length(files)) {
  stop("Each file in data/listings/ must be a different month")
}

table(chch$month_date)
dim(chch)
```

# 4. Save

```{r save}
write.csv(chch, "output/Christchurch_combined.csv", row.names = FALSE)
```
